---
title: "Day 1 & 2: Conceptual practice exercises"
subtitle: "Lines, rates of change, functions, exponentials & logs, and integrals"
editor_options:
  chunk_output_type: console
---

::: {.callout-note}
For every exercise below, write your final answer as a **complete sentence**, including units, as if you were explaining it to someone unfamiliar with the data. Where relevant, avoid overstating certainty (e.g. prefer "increased by an average of..." over "increased by...").
:::

## Exercise 1

A stream's water temperature increases as you move downstream from a shaded headwater spring. At a monitoring station 2 km downstream, the water temperature is 14°C. At a station 5 km downstream, the temperature is 20°C. Assume temperature increases linearly with distance downstream.

(a) Find the equation for water temperature $T$ (°C) as a function of distance downstream $x$ (km).
(b) Create a plot showing the line. Clearly label the axes, including units, and add a title . 
(c) What is the $y$-intercept of this line, and what does it mean in the context of the stream? Answer in a full sentence.

::: {.callout-tip collapse="true"}
## Solution

**(a)** Using the two points $(2, 14)$ and $(5, 20)$:

$$
m = \frac{20-14}{5-2} = \frac{6}{3} = 2
$$

Using point-slope form with $(2,14)$:
$$
T - 14 = 2(x-2) \ \Rightarrow \ T = 2x + 10
$$

**(b)**

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
x <- seq(0, 8, length.out = 100)
T <- 2*x + 10

plot(x, T, type = "l", lwd = 3, col = "green",
     xlim = c(0, 8), ylim = c(0, 24),
     xlab = "Distance downstream (km)", ylab = "Water temperature (°C)",
     xaxt = "n", yaxt = "n")
abline(h = seq(0, 24, by = 2), col = "lightgray", lty = "dotted")
abline(v = seq(0, 8, by = 1), col = "lightgray", lty = "dotted")
axis(1, at = seq(0, 8, by = 1))
axis(2, at = seq(0, 24, by = 2))
points(c(2,5), c(14,20), pch = 19, cex = 1.3, col = "blue")
lines(x, T, lwd = 2)
```

**(c)** The $y$-intercept is $b = 10$.

> "According to this model, the water at the very start of the stream (right at the headwater spring, 0 km downstream) is predicted to be 10°C."
:::

## Exercise 2

Researchers plotted [dissolved oxygen (DO)](https://www.epa.gov/system/files/documents/2021-07/parameter-factsheet_do.pdf) concentration against water depth in a lake, and found the relationship was linear over the depth range sampled.

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
depth <- seq(0, 12, length.out = 100)
DO <- -0.6*depth + 9

plot(depth, DO, type = "l", lwd = 3, col = "green",
     xlim = c(0, 12), ylim = c(0, 10),
     xlab = "Depth (m)", ylab = "Dissolved oxygen (mg/L)",
     xaxt = "n", yaxt = "n")
abline(h = seq(0, 10, by = 1), col = "lightgray", lty = "dotted")
abline(v = seq(0, 12, by = 1), col = "lightgray", lty = "dotted")
axis(1, at = seq(0, 12, by = 1))
axis(2, at = seq(0, 10, by = 1))
lines(depth, DO, lwd = 2)
```

(a) Write the equation for dissolved oxygen, $DO$, as a function of depth.
(b) Interpret the slope in a full sentence, including units.

::: {.callout-tip collapse="true"}
## Solution

**(a)** The line passes through $(0, 9)$ and $(10, 3)$, so:

- $y$-intercept: $9$
- slope: $\frac{3-9}{10-0} = \frac{-6}{10} = -0.6$

$DO(depth) = -0.6 \cdot depth + 9$

**(b)**

> "For every additional meter of depth, dissolved oxygen concentration decreases by an average of 0.6 mg/L, starting from a surface concentration of about 9 mg/L."
:::

## Exercise 3

Right after an accidental spill, ecologists tracked the concentration of a pollutant, $C(t)$ (mg/L), in a pond over the following 20 days, where $t$ is days since the spill.

:::: {.columns}
::: {.column width="35%"}
| $t$ (days) | $C(t)$ (mg/L) |
|:---:|:---:|
| 2  | 14.4 |
| 6  | 33.6 |
| 10 | 40.0 |
| 14 | 33.6 |
| 18 | 14.4 |
:::

::: {.column width="65%"}
```{r}
#| echo: false
#| fig-align: center
#| out-width: "95%"
t <- seq(0, 20, length.out = 200)
C <- 8*t - 0.4*t^2

plot(t, C, type = "l", lwd = 3, col = "#003660",
     xlim = c(0, 20), ylim = c(0, 45),
     xlab = "Days since spill", ylab = "Pollutant concentration (mg/L)")
points(c(2,6), c(8*2-0.4*2^2, 8*6-0.4*6^2), pch = 19, cex = 1.4, col = "#FEBC11")
segments(2, 8*2-0.4*2^2, 6, 8*6-0.4*6^2, col = "#FEBC11", lwd = 2, lty = 2)
```
:::
::::

(a) Calculate the average rate of change in pollutant concentration between day 2 and day 6.
(b) Write your answer as a full sentence, with units, without overstating certainty.

::: {.callout-tip collapse="true"}
## Solution

$$
\text{average rate of change} = \frac{C(6) - C(2)}{6-2} = \frac{33.6 - 14.4}{4} = \frac{19.2}{4} = 4.8 \ \frac{\text{mg/L}}{\text{day}}
$$

> "Between day 2 and day 6 after the spill, the pollutant concentration increased by an average of 4.8 mg/L per day."
:::

## Exercise 4: Average vs. instantaneous rate of change

Using the same pollutant concentration curve $C(t) = 8t - 0.4t^2$ from Exercise 3, let's zoom in on what's happening right at $t=10$, where the curve reaches its peak.

Fill in the table below, computing the *average* rate of change of $C$ between $t=10$ and $t = 10+\Delta x$, for smaller and smaller values of $\Delta x$:

| $\Delta x$ | $t=10+\Delta x$ | $C(10+\Delta x)$ | average rate of change on $[10, 10+\Delta x]$ |
|:---:|:---:|:---:|:---:|
| 4    | 14   | 33.6  |  |
| 2    | 12   | 38.4  |  |
| 1    | 11   | 39.6  |  |
| 0.5  | 10.5 | 39.9  |  |

(a) As $\Delta x$ gets smaller, what value does the average rate of change appear to be approaching?
(b) What do we call this converging value, and what does it tell us about the pollutant concentration at *exactly* $t=10$? Answer in a full sentence.
(c) Conceptually, based on the shape of the graph, why does it make sense that the instantaneous rate of change takes this value at $t=10$?

::: {.callout-tip collapse="true"}
## Solution

| $\Delta x$ | $t=10+\Delta x$ | $C(10+\Delta x)$ | average rate of change on $[10, 10+\Delta x]$ |
|:---:|:---:|:---:|:---:|
| 4    | 14   | 33.6  | $-1.6$ |
| 2    | 12   | 38.4  | $-0.8$ |
| 1    | 11   | 39.6  | $-0.4$ |
| 0.5  | 10.5 | 39.9  | $-0.2$ |

**(a)** The average rate of change appears to be converging to $0$ as $\Delta x \to 0$.

**(b)** This converging value is the **instantaneous rate of change** (the derivative) of $C$ at $t=10$.

> "At exactly day 10 after the spill, the pollutant concentration is momentarily neither increasing nor decreasing."

**(c)** Day 10 is the peak of the curve — the concentration stops rising and is about to start falling. At a peak (or any turning point), the tangent line to the curve is flat, so the instantaneous rate of change is zero, even though the *average* rate of change over a wider interval around that point would generally not be exactly zero.
:::

## Exercise 5: Translating a scenario into function notation

Environmental scientists model evapotranspiration (the combined loss of water from soil evaporation and plant transpiration) for a crop field. They believe evapotranspiration depends on: solar radiation, air temperature, and wind speed.

(a) In this scenario, which variable is the output (dependent variable), and which are the inputs (independent variables)?
(b) Write a sentence of the form "[output] is a function of [input(s)]" using the variable names.
(c) Choose short notation (a letter or symbol) and units for each variable.
(d) Using your notation, write the function notation set-up for this relationship (you do not need an explicit formula).
(e) Suppose a simplified model turns out to be:
$$
ET(R,T,W) = 0.02R + 0.15T - 0.05W
$$
where $ET$ is evapotranspiration (mm/day), $R$ is solar radiation (W/m²), $T$ is air temperature (°C), and $W$ is wind speed (km/hr). Evaluate this model when $R=500$, $T=25$, and $W=10$, and state the result as a full sentence.

::: {.callout-tip collapse="true"}
## Solution

**(a)** Evapotranspiration is the output (dependent variable). Solar radiation, air temperature, and wind speed are the inputs (independent variables).

**(b)** For example: $ET$ = evapotranspiration (mm/day), $R$ = solar radiation (W/m²), $T$ = air temperature (°C), $W$ = wind speed (km/hr). *(Any reasonable, clearly-defined notation is fine!)*

**(c)** $$ET(R, T, W) = [\text{expression containing } R, T, \text{ and } W]$$

**(d)** "Evapotranspiration is a function of solar radiation, air temperature, and wind speed."

**(e)**
$$
ET(500, 25, 10) = 0.02(500) + 0.15(25) - 0.05(10) = 10 + 3.75 - 0.5 = 13.25
$$

> "When solar radiation is 500 W/m², air temperature is 25°C, and wind speed is 10 km/hr, the model predicts an evapotranspiration rate of 13.25 mm per day."
:::

## Exercise 6: Exponential growth and solving with logarithms

An algal bloom is discovered in a lake. Researchers estimate the algae population grows exponentially, following $N(t) = N_0 e^{rt}$, where $N_0$ is the initial cell count and $t$ is time in days since discovery.

At discovery ($t=0$), the population is $N_0 = 150$ cells/mL. Four days later, the population has grown to $400$ cells/mL.

(a) Solve for the growth rate $r$.
(b) Lake managers will close the lake to recreation if the algae population reaches $5{,}000$ cells/mL. Using your value of $r$, solve for the time $t$ at which this threshold will be reached.
(c) State your answer to (b) as a full sentence.

::: {.callout-tip collapse="true"}
## Solution

**(a)**
$$
\begin{aligned}
400 &= 150e^{4r} \\
\frac{400}{150} &= e^{4r} \\
\ln\!\left(\frac{400}{150}\right) &= 4r \\
r &= \frac{\ln(2.667)}{4} \approx 0.245
\end{aligned}
$$

**(b)**
$$
\begin{aligned}
5000 &= 150e^{0.245t} \\
\frac{5000}{150} &= e^{0.245t} \\
\ln(33.33) &= 0.245t \\
t &= \frac{\ln(33.33)}{0.245} \approx 14.3
\end{aligned}
$$

**(c)**

> "If the current growth trend continues, the algal population is expected to reach the 5,000 cells/mL closure threshold about 14.3 days after it was first discovered."
:::

## Exercise 7: Integrals, conceptually

After restoring a section of coastal wetland, ecologists measure the rate at which the wetland absorbs stormwater runoff, $r(t)$ (in m³/hour), throughout a 24-hour storm. The graph below shows this absorption rate over time.

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
t <- seq(0, 24, length.out = 200)
rate <- 200 * exp(-((t-9)^2)/40)

plot(t, rate, type = "l", lwd = 3, col = "#003660",
     xlim = c(0,24), ylim = c(0,220),
     xlab = "Hours since storm began", ylab = "Stormwater absorption rate (m³/hr)")
abline(v = seq(0,24, by = 6), col = "lightgray", lty = "dotted")
```

(a) What does the *area under this curve* represent, in words? Be sure to include units.
(b) Looking at the shape of the graph (without calculating anything), during which 6-hour window — hours 0–6, 6–12, 12–18, or 18–24 — do you expect the *largest volume* of stormwater to have been absorbed? Explain your reasoning.
(c) Suppose the wetland absorbed water at a roughly constant rate of 150 m³/hour between hour 6 and hour 8. Estimate the volume of water absorbed during that window, and state your answer as a full sentence.
(d) In your own words, explain why we *can't* just take the absorption rate at a single instant in time and multiply it by 24 hours to estimate the total volume absorbed over the whole storm — the way we could if the rate had stayed constant.

::: {.callout-tip collapse="true"}
## Solution

**(a)** The area under the rate curve represents the **total volume of stormwater absorbed by the wetland** over that time period, in m³ (rate in m³/hr $\times$ time in hr = volume in m³).

**(b)** Hours 6–12. The curve is highest and covers the most area in that window, since the absorption rate peaks around hour 9 — the middle of the storm, when the ground is saturated and inflow is greatest.

**(c)**
$$
150\ \tfrac{\text{m}^3}{\text{hr}} \times 2\ \text{hr} = 300\ \text{m}^3
$$

> "Between hour 6 and hour 8 of the storm, the wetland absorbed an estimated 300 m³ of stormwater."

**(d)** The absorption rate isn't constant — it changes throughout the storm (rising, peaking, then falling). Multiplying a single instantaneous rate by 24 hours would only be accurate if that rate held steady the whole time, which it clearly doesn't (you'd either over- or under-estimate the total, depending on which instant you picked). Because the rate varies continuously, we need to add up (integrate) the contribution of every instant across the storm to get the true total volume — that's exactly what "area under the curve" captures.
:::
